---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r initialize,echo=FALSE}
library(lattice)
options(scipen = 10)
```

***


## Loading and preprocessing the data  


```{r loaddata}
file <- unz("activity.zip", "activity.csv")
mydata <- read.csv(file, na.strings = "NA", comment.char = "",
                   colClasses = c("numeric", "character", "numeric"))

## process and transform data
mydata <- within(mydata, {
    date <- as.Date(date)   # coerce date
    
    # convert interval to elapsed minutes since midnight
    interval <- interval - (interval %/% 100) * 40

    # add day to indicate whether weekday/weekend
    day <- factor(weekdays(date) %in% c("Saturday", "Sunday"), 
                  levels = c(T, F), labels = c("Weekend", "Weekday"))
})

head(mydata)
```

*__Interval__ is converted to minutes elapsed since midnight 
on the corresponding day to improve graph accuracy*  
*New column __Day__ is added to indicate whether the day 
is a weekday or weekend*


***


## Total number of steps taken per day  


### Estimates

```{r totalsteps}
totalStepsPerDay <- aggregate(steps ~ date, mydata, sum)$steps
mean.totalStepsPerDay <- mean(totalStepsPerDay)
median.totalStepsPerDay <- median(totalStepsPerDay)
```

Mean of total steps per day is __`r round(mean.totalStepsPerDay, 2)`__  
Median of total steps day is __`r median.totalStepsPerDay`__


### Distribution

```{r totalstepsHistBeforeImputing,fig.width=5,fig.height=5}
hist(totalStepsPerDay, breaks = 10, col = "salmon", 
     xlab = "Total steps per day", main = NULL)
```


***


## Average daily activity pattern


### Highest activity interval

```{r activityInterval}
activity <- aggregate(steps ~ interval, mydata, mean)
maxActivityInterval <- with(activity, interval[which.max(steps)])
maxActivityInterval
```

*Note: Units in mins after midnight*  
 

### Time series plot

```{r activityTimeSeriesPlot}
plot(activity, type = "l", xlab = "Interval", ylab = "Steps", 
     main = "Average daily activity pattern")
```


***


## Imputing missing values


```{r missingVals}
nrMissingVals <- sum(is.na(mydata$steps))
```

Number of missing values in the dataset = __`r nrMissingVals`__


### Strategy

The NAs in the dataset are first replaced with the mean steps for that day.
If there exists any remaining NAs they are replaced with the mean steps for
that interval. The second step handles a *very odd scenario* in which
no values exist on a particular day, i.e. its mean steps = NA on that day.

```{r imputeStrategy}
imputedData <- within(mydata, {
    # replace NAs with mean steps for that day
    steps <- tapply(steps, date, 
                    function(x) { x[is.na(x)] <- mean(x, na.rm = T); x })
    steps <- unsplit(steps, date)
    
    # replace remaining NAs with mean steps for that interval
    steps <- tapply(steps, interval, 
                    function(x) { x[is.na(x)] <- mean(x, na.rm = T); x })
    steps <- unsplit(steps, interval)
})
```


### Impact on total steps taken per day


#### Estimates

```{r totalStepsAfterImputing}
totalStepsPerDay.imputed <- aggregate(steps ~ date, imputedData, sum)$steps
mean.totalStepsPerDay.imputed <- mean(totalStepsPerDay.imputed)
median.totalStepsPerDay.imputed <- median(totalStepsPerDay.imputed)
```

Mean total steps per day is __`r round(mean.totalStepsPerDay.imputed, 2)`__  
Median total steps per day is __`r round(median.totalStepsPerDay.imputed, 2)`__  


#### Distribution

```{r totalStepsHistAfterImputing,fig.width=5,fig.height=5}
hist(totalStepsPerDay.imputed, breaks = 10, col = "wheat",
     xlab = "Total steps per day", main = NULL)
```


#### Summary statistics comparison

```{r imputeCompareSummary}
summary(totalStepsPerDay, digits = 12)      # before imputing
summary(totalStepsPerDay.imputed, digits = 12)      # after imputing
```


***


## Activity patterns between weekdays and weekends

```{r activityWeekdayWeekend,fig.width=9,fig.width=9}
timeSeriesData <- aggregate(steps ~ interval + day, imputedData, mean)
xyplot(steps ~ interval | day, data = timeSeriesData, type = "l",
       layout = c(1, 2), xlab = "Interval", ylab = "Steps")
```    
       
       
       
       